Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-07-28 article.tex
after professional editor corrections

\vspace{-0.8cm}
\begin{abstract}
In this work, we propose a simple, but yet efficient method for the problem of
connectome inference in calcium imaging data. The proposed algorithm consists of
two steps. First, processing the raw signals to detect neural peak activities.
Second, inferring the degree of association between neurons from partial
correlation statistics. This paper summarizes the methodology that led us to
win the Connectomics Challenge, proposes a simplified version of our method and
finally discusses our results with respect to other inference methods.
\end{abstract}

@@

The human brain is a complex biological
organ made of about 100 billion of neurons, each connected to 7000 other neurons on
average \citep{pakkenberg2003aging}. Unfortunately, direct observation of the connectome, the wiring
diagram of the brain, is not yet technically feasible. Without being perfect,
calcium imaging currently allows the real-time and simultaneous observation of
neuron activities from thousands of neurons, producing individual time series
representing their fluorescence intensity. From these data, the connectome
inference problem amounts to retrieving the synaptic connections between
neurons on the basis of the fluorescence time series. This
problem is often difficult because of experimental issues, including
masking effects (i.e., some of the neurons are not observed or confounded with others), the low sampling rate of the optical device with respect to the
neural activity speed or the slow decay of fluorescence.

Formally, the connectome can be represented as a directed graph $G=(V,E)$,
where $V$ is a set of $p$ nodes, representing neurons, and $E \subseteq
\left\{(i, j) \in V \times V\right\}$ is a set of edges, representing direct
synaptic connections between neurons. Causal interactions are expressed by the
direction of edges: $(i, j) \in E$ indicates that the state of neuron $j$ might
be caused by the activity of neuron $i$. In those terms,  the connectome
inference problem is formally stated as follows:  \textit{Given the sampled
observations $\{ x^t_i \in \mathbb{R} | i \in V, t = 1, \dots, T \}$ of $p$
neurons for $T$ time intervals, the goal is to infer the set $E$ of connections
in $G$.}

In this paper, we present a simplified -- yet nearly as good -- version of the
winning method\footnote{Code available at \url{https://github.com/asutera/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The paper is structured as follows:
Section~\ref{sec:filter} describes the signal processing methods applied on
fluorescent calcium time series; Section \ref{sec:inference} then presents the
proposed approach and its theoretical properties; Section~\ref{sec:results}
provides an empirical analysis and comparison with other network inference
methods, while Section~\ref{sec:conclusion} finally discusses our work and
provides further research directions. Additionally,
Appendix~\ref{app:optimized} further describes, in all details, our actual
winning method, giving slightly better results than the method presented in
the paper, at the price of parameter tuning. Appendix~\ref{app:supp} provides supplementary results on other datasets.

@@

Under the simplifying assumption that neurons are on-off units, characterized
by short periods of intense activities, or peaks, and longer periods of
inactivity, the first part of our algorithm consists in cleaning the raw
fluorescence data.
More specifically, time series are processed using standard
signal processing filters in order to (i) remove noise mainly due to fluctuations independent of calcium, calcium fluctuations independent of spiking activity, calcium fluctuations in nearby tissues that have been mistakenly captured, or simply by the imaging process (ii) to account for fluorescence low decay and (iii) to reduce the importance of
global high activity in the network. The overall process is illustrated in
Figure~\ref{fig:filtered-signal}.

@@

As Figure~\ref{fig:original_curve} shows, the raw
fluorescence signal is very noisy due to light scattering artifacts that
usually affect the quality of the recording~\citep{lichtman2011big}.
Accordingly, the first step of our pipeline is to smoothen the signal, using
one of the following low-pass filters for filtering out high frequency noise:
\begin{align}
f_1(x^t_i) &= x^{t-1}_i + x^t_i + x^{t+1}_i, \label{eq:symetric-median} \\
f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_i^t.
\label{eq:weighted-asymetric-median}
\end{align}
These filters are standard in the signal processing field \citep{kaiser1977data, oppenheim1983signals}. For the sake of illustration, the effect of the filter $f_1$ on the signal
is shown in Figure \ref{fig:lp_curve}.

Furthermore, short spikes, characterized by a high
frequency, can be seen as an indirect indicator of neuron communication while low frequencies of the signal mainly correspond to the slow
decay of fluorescence. To have a signal that only has high magnitude around instances where the spikes occur, the second step of our pipeline transforms the time series into its backward
difference
\begin{align}
g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i, \label{eq:high-pass-filter}
\end{align}
as shown in Figure \ref{fig:hp_curve}.

To filter out small variations in the signal obtained after applying the
function $g$ as well as to eliminate negative values, we use the following
hard-threshold filter
\begin{align}\label{eqn:hfilter}
h(x^{t}_i) &= x^{t}_i \mathbb{1}(x^{t}_i \geq \tau) \text{ with } \tau > 0,
\end{align}
yielding Figure \ref{fig:threshold_curve} where $\tau$ is the threshold parameter and $\mathbb{1}$ is the indicator function.
As we can see, the processed signal only contains clean spikes.

The objective of the last step of our filtering procedure is to decrease the
importance of spikes that occur when there is a high global activity in the
network with respect to  spikes that occur during normal activity. Indeed, we
have conjectured that when a large part of the network is firing, the rate at
which observations are made is not high enough for being able to detect
interactions, and that it would therefore be preferable to lower their
importance by changing appropriately their magnitude. Additionally, it is
well-known that neurons may also spike because of a global high activity
\citep{stetter2012model}. In such  context, detecting pairwise neuron
interactions from the firing activity is meaningless. As such,
the signal output by $h$ is finally applied to the following function
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}}, \label{eq:magnify-filter}
\end{align}
whose effect is to magnify the importance of spikes that occur in case of low
global activity (measured by $\sum_{j} x^{t}_j$), as observed for instance
around $t=4\text{s}$ in Figure~\ref{fig:weight_curve}. Note that the particular case where there
is no activity, i.e., $\sum_{j} x^{t}_j = 0$, is solved by setting $w(x^{t}_i)
= 1$.

To summarize, the full signal processing pipeline of our simplified approach is defined by the composed function $w \circ h \circ g \circ
f_1$ (resp. $f_2$). When applied to the raw signal of Figure
\ref{fig:original_curve}, it outputs the signal shown in Figure
\ref{fig:weight_curve}.

@@

Our procedure to infer connections between neurons first assumes that
the (filtered) fluorescence concentrations of all $p$ neurons at each
time point can be modeled as a set of random variables $X = \{X_1,
\dots, X_p\}$ that are independently drawn from the same time invariant
joint probability distribution $P_X$. 
As a consequence, our inference method does not exploit the time
ordering of the observations (although time ordering is exploited by
the filters).

Given this assumption, we then propose to use as a measure of the
strength of the connection between two neurons $i$ and $j$, the
\textit{Partial correlation} coefficient $p_{i,j}$ between their corresponding
random variables $X_i$ and $X_j$, defined by:
\begin{equation}
p_{i,j} =
-\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}, \label{eq:inverse}
\end{equation}
where $\Sigma^{-1}$, known as the precision or concentration matrix, is the inverse of the covariance matrix $\Sigma$ of $X$. 
Assuming that the distribution $P_X$ is a multivariate Gaussian
distribution ${\cal N}(\mu,\Sigma)$, it can be shown that $p_{i,j}$ is
zero if and only if $X_i$ and $X_j$ are independent given all other
variables in $X$, i.e., $X_i \perp X_j|X^{-i,j}$ where $X^{-i,j}= X
\setminus\{X_i,X_j\}$. Partial correlation thus measures conditional
dependencies between variables. It should thus naturally only detect direct associations
between neurons and filter out spurious indirect effects. The interest
of partial correlation as an association measure has already been
shown for the inference of gene regulatory networks
\citep{de2004discovery,Schafer:2005}.
Note that the partial correlation statistic is symmetric
(i.e. $p_{i,j}=p_{j,i}$). Therefore, our approach cannot identify the
direction of the interactions between neurons. We will see in
Section~\ref{sec:results} why this only affects slightly its
performance, with respect to the metric used in the Connectomics
challenge.

Practically, the computation of all $p_{i,j}$ coefficients using Equation
\ref{eq:inverse} requires the estimation of the covariance matrix $\Sigma$
and then computing its inverse. Given that we have typically more
samples than neurons, the covariance matrix can be inverted in a
straightforward way. We nevertheless obtained some improvement by
replacing the exact inverse with an approximation using only the $M$
first principal components \citep{bishop2006pattern} (with
$M=0.8 p$ in our experiments, see Appendix~\ref{app:pca}).

Finally, let us note that the performance of our simple method appears to
be quite sensitive to the values of parameters (e.g., choice of $f_1$ or $f_2$
or the value of the threshold $\tau$) in the combined function of the
filtering and inferring processes. One approach, further referred
to as \textit{Averaged Partial correlation} statistics, for improving
its robustness is to average correlation statistics over various
values of the parameters, thereby reducing the variance of its
predictions. Further details about parameter selection are provided in
Appendix~\ref{app:optimized}.

@@

We report here experiments on the \textit{normal-1,2,3}, and \textit{4}
datasets provided by the organizers of the Connectomics challenge (see
Appendix~\ref{app:supp} for experiments on other datasets). Each of
these datasets is obtained from the simulation \citep{stetter2012model} of
different neural networks of 1000 neurons and about 15000 edges each (i.e., a
network density of about 1.5\%). Each neuron is described by a calcium
fluorescence time series of length $T=179500$. All inference methods compared
here provide a ranking of all pairs of neurons according to some association score. To assess the quality of this ranking, we compute both ROC and
precision-recall curves against the ground truth network, which are summarized
using the area under the curve and respectively  denoted AUROC and AUPRC. Only
the AUROC score was used to rank the challenge participants but the precision-recall curve has been shown to be a more sensible metric for network
inference, especially when network density is small (see e.g.,
\cite{schrynemackers2013protocols}). Since neurons are not self-connected in
the ground truth networks (i.e., $(i, i) \not \in E, \forall i \in V$), we
have manually set the score of such edges to the minimal possible association
score before computing ROC and PR curves.

@@

The top of Table \ref{tab:comparison} reports AUROC and AUPRC for all
four networks using in each case partial correlation with different
filtering functions. Except for the last two rows that use PCA, the
exact inverse of the covariance matrix was used in each case. These
results clearly show the importance of the filters. AUROC increases in
average from 0.77 to 0.93. PCA does not really affect AUROC scores but
it significantly improves AUPRC scores. Taking the average over
various parameter settings gives an improvement of 10\% in AUPRC but
only a minor change in AUROC. The last row (``Full method'') shows the
final performance of the method specifically tuned for the challenge
(see Appendix \ref{app:optimized} for all details). Although this
tuning was decisive to obtain the best performance at the challenge,
it does not significantly improve neither AUROC nor AUPRC.

@@

In the bottom of Table \ref{tab:comparison}, we provide as a comparison the
performance of three other methods: standard (Pearson) correlation (PC),
generalized transfer entropy (GTE), and GENIE3. ROC and PR curves on the
\textit{normal-2} network are shown for all methods in Figure
\ref{fig:curves}. Pearson correlation measures the unconditional linear
(in)dependence between variables and it should thus not be able to filter out
indirect interactions between neurons. GTE \citep{stetter2012model} was
proposed as a baseline for the challenge. This method builds on Transfer
Entropy to measure the association between two neurons. Unlike our approach, it
can predict the direction of the edges. GENIE3 \citep{huynhthu2010inferring} is
a gene regulatory network inference method that was the best performer of the
DREAM5 challenge \citep{marbach2012}. When transposed to neural networks, this
method uses as a confidence score for the edge going from neuron $i$ to neuron
$j$ the importance score of variable $X_i$ in a Random Forest model trying to
predict $X_j$ from all variables in $X\setminus X_j$. To reduce the
computational cost of this method, we had however to limit each tree in the
Random Forest model to a maximum depth of 3. This constraint potentially
severely affects the performance of this method with respect to the use of
fully grown trees. PC and GENIE3 were applied on the time series filtered using the functions $w\circ h\circ g$ and $h\circ g\circ f_1$ (which
gave the best performance), respectively. For GENIE3, we built 10000 trees per neuron and we
used default settings for all other parameters (except for the maximal tree
depth). For GTE, we reproduced the exact same setting (conditioning level and
pre-processing) that was used by the organizers of the challenge.

Partial correlation and averaged partial correlation clearly outperform all
other methods on all datasets (see Table \ref{tab:comparison} and Appendix \ref{app:supp}). The
improvement is more important in terms of AUPRC than in terms of AUROC. As
expected, Pearson correlation performs very poorly in terms of AUPRC. GTE and
GENIE3 work much better but these two methods are nevertheless clearly below
partial correlation. Among these two methods, GTE is slightly better in terms
of AUROC, while GENIE3 is significantly better in terms of AUPRC. Given that we
had to limit this latter method for computational reasons, these results are
very promising and a comparison with the full GENIE3 approach is certainly part
of our future works.

The fact that our method is unable to predict edge directions does not seem to
be a disadvantage with respect to GTE and GENIE3. Although partial correlation
scores each edge and its opposite similarly, it can reach precision values
higher than 0.5 (see Figure \ref{fig:curves}(b)), suggesting that it mainly ranks high
pairs of neurons that interact in both directions.  It is interesting also to
note that, on \textit{normal-2}, a method that perfectly predicts the
undirected network (i.e., that gives a score of $1$ to each pair $(i,j)$ such that
$(i,j)\in E$ or $(j,i)\in E$, and $0$ otherwise) already reaches an AUROC as high
as $0.995$ and an AUPRC of $0.789$.

@@

In this paper, we outlined a simple but efficient methodology for the problem
of connectome inference from calcium imaging data. Our approach consists of two
steps: (i) processing fluorescence data to detect neural peak activities and
(ii) inferring the degree of association between neurons from partial
correlation statistics. Its simplified variant outperforms other
network inference methods while its optimized version proved to be the best method
on the Connectomics challenge. Given its simplicity and good performance, we
therefore believe that the methodology presented in this work
would constitute a solid and easily reproducible baseline for further work in
the field of connectome inference.

\paragraph{Acknowledgments}
A. Joly and G. Louppe are research fellows of the FNRS, Belgium.  A. Sutera is
recipient of a FRIA fellowship of FRS-FNRS, Belgium. This work is supported by
PASCAL2 and the IUAP DYSCO, initiated by the Belgian State, Science Policy
Office.

@@

This section provides a detailed description of the method specifically tuned
for the Connectomics challenge. We restrict our description to the
differences with respect to the simplified method presented in the main
paper. Most parameters were tuned so as to maximize AUROC on the
\textit{normal-1} dataset and our design choices were validated by monitoring
the AUROC obtained by the 145 entries we submitted during the
challenge. Although the tuned method performs better than the simplified one on
the challenge dataset, we believe that the tuned method clearly overfits the
simulator used to generate the challenge data and that the simplified method
should work equally well on new independent datasets. We nevertheless provide
the tuned method here for reference purpose. Our implementation of the tuned
method is available at \url{https://github.com/asutera/kaggle-connectomics}.

@@

\begin{itemize}
\item In addition to $f_1$ and $f_2$ (Equations \ref{eq:symetric-median} and
  \ref{eq:weighted-asymetric-median}), two alternative low-pass filters $f_3$
  and $f_4$ are considered:
\begin{align}
f_3(x^t_i) &= x^{t-1}_i + x^{t}_i + x^{t+1}_i + x^{t+2}_i, \label{eq:asymetric-median-forward} \\
f_4(x^t_i) &=  x_i^t + x^{t+1}_i  + x^{t+2}_i + x^{t+3}_i. \label{eq:asymetric-median}
\end{align}
\item An additional filter $r$ is applied to smoothen differences in peak magnitudes
  that might remain after the application of the hard-threshold filter $h$:
\begin{align}
r(x^t_i) = (x_i^t)^c,
\end{align}
with $c=0.9$.
\item Filter $w$ is replaced by a more complex filter $w^*$ defined as:
\begin{align}
 w^*(x^{t}_i) &= {(x^{t}_i + 1 )^{\left (1 + \frac{1}{\sum_{j} x^{t}_j}\right )}}^{k(\sum_{j} x^{t}_j)}
\end{align}
where the function $k$ is a piecewise linear function optimized separately for
each filter $f_1$, $f_2$, $f_3$ and $f_4$ (see the implementation for full
details). Filter $w$ in the simplified method is a special case of $w^*$ with
$k(\sum_j x_j^t)=1$.
\end{itemize}
The pre-processed time-series are then obtained by the application of the
following function: $w^*\circ r \circ h \circ g \circ f_i$ (with $i=1$, 2, 3, or 4).

@@

As discussed in Section \ref{sec:inference}, the performance of the method (in
terms of AUROC) is sensitive to the value of the parameter $\tau$ of the
hard-threshold filter $h$ (see Equation \ref{eqn:hfilter}), and to the choice
of the low-pass filter (among $\{f_1, f_2, f_3, f_4\}$).
Like in the simplified method, we have averaged the partial correlation statistics obtained for all the pairs $(\tau,\mbox{low-pass filter}) \in \{0.100,0.101,\ldots,0.209\}\times \{f_1, f_2, f_3, f_4\}$.

Filters $f_1$ and $f_2$ display similar performances and thus were given similar
weights (i.e., resp. $0.383$ and $0.345$). These weights were chosen equal to the weights selected for the simplified method. In contrast, filters $f_3$
and $f_4$ turn out to be individually less competitive and were therefore given
smaller importance in the weighted average (i.e., resp. $0.004$ and $0.268$). Yet, as further shown in
Section~\ref{sapp:results}, combining all $4$ filters proves to slightly
improve performance with respect to using $f_1$ and $f_2$ only.

>>>>

\vspace{-0.8cm}
\begin{abstract}
In this work, we propose a simple yet effective solution to the problem of
connectome inference in calcium imaging data. The proposed algorithm consists of
two steps. First, processing the raw signals to detect neural peak activities.
Second, inferring the degree of association between neurons from partial
correlation statistics. This paper summarises the methodology that led us to
win the Connectomics Challenge, proposes a simplified version of our method, and
finally compares our results with respect to other inference methods.
\end{abstract}

@@

The human brain is a complex biological organ made of about 100 billion of
neurons, each connected to, on average, 7,000 other neurons
\citep{pakkenberg2003aging}. Unfortunately, direct observation of the
connectome, the wiring diagram of the brain, is not yet technically feasible.
Without being perfect, calcium imaging currently allows for real-time and
simultaneous observation of neuron activity from thousands of neurons,
producing individual time-series representing their fluorescence intensity.
From these data, the connectome inference problem amounts to retrieving the
synaptic connections between neurons on the basis of the fluorescence time-
series. This problem is difficult to solve because of experimental issues,
including masking effects (i.e., some of the neurons are not observed or
confounded with others), the low sampling rate of the optical device with
respect to the neural activity speed, or the slow decay of fluorescence.

Formally, the connectome can be represented as a directed graph $G=(V,E)$,
where $V$ is a set of $p$ nodes representing neurons, and $E \subseteq
\left\{(i, j) \in V \times V\right\}$ is a set of edges representing direct
synaptic connections between neurons. Causal interactions are expressed by the
direction of edges: $(i, j) \in E$ indicates that the state of neuron $j$ might
be caused by the activity of neuron $i$. In those terms,  the connectome
inference problem is formally stated as follows:  \textit{Given the sampled
observations $\{ x^t_i \in \mathbb{R} | i \in V, t = 1, \dots, T \}$ of $p$
neurons for $T$ time intervals, the goal is to infer the set $E$ of connections
in $G$.}

In this paper, we present a simplified - and almost as good - version of the
winning method\footnote{Code available at \url{https://github.com/asutera/kaggle-connectomics}} of the Connectomics
Challenge\footnote{\url{http://connectomics.chalearn.org}}, as a simple and
theoretically grounded approach based on signal processing techniques and
partial correlation statistics. The paper is structured as follows:
Section~\ref{sec:filter} describes the signal processing methods applied on
fluorescent calcium time series; Section\ref{sec:inference} then presents the
proposed approach and its theoretical properties; Section~\ref{sec:results}
provides an empirical analysis and comparison with other network inference
methods, while finally, in Section~\ref{sec:conclusion} we discuss our work and
provide further research directions. Additionally,
Appendix~\ref{app:optimized} further describes, in full detail, our actual
winning method which gives slightly better results than the method presented in
this paper, at the cost of parameter tuning. Appendix~\ref{app:supp} provides supplementary results on other datasets.

@@

Under the simplifying assumption that neurons are on-off units, characterised
by short periods of intense activity, or peaks, and longer periods of
inactivity, the first part of our algorithm consists of cleaning the raw
fluorescence data.
More specifically, time series are processed using standard
signal processing filters in order to : (i) remove noise mainly due to fluctuations independent of calcium, calcium fluctuations independent of spiking activity, calcium fluctuations in nearby tissues that have been mistakenly captured, or simply by the imaging process ; (ii) to account for fluorescence low decay ; and (iii) to reduce the importance of
high global activity in the network. The overall process is illustrated in
Figure~\ref{fig:filtered-signal}.

@@

As Figure~\ref{fig:original_curve} shows, the raw
fluorescence signal is very noisy due to light scattering artifacts that
usually affect the quality of the recording~\citep{lichtman2011big}.
Accordingly, the first step of our pipeline is to smoothe the signal, using
one of the following low-pass filters for filtering out high frequency noise:
\begin{align}
f_1(x^t_i) &= x^{t-1}_i + x^t_i + x^{t+1}_i, \label{eq:symetric-median} \\
f_2(x^t_i) &= 0.4 x^{t-3}_i + 0.6 x^{t-2}_i + 0.8 x^{t-1}_i + x_i^t.
\label{eq:weighted-asymetric-median}
\end{align}
These filters are standard in the signal processing field \citep{kaiser1977data, oppenheim1983signals}. For the purposes of illustration, the effect of the filter $f_1$ on the signal
is shown in Figure \ref{fig:lp_curve}.

Furthermore, short spikes, characterized by a high
frequency, can be seen as an indirect indicator of neuron communication, while low frequencies of the signal mainly correspond to the slow
decay of fluorescence. To have a signal that only has high magnitude around instances where the spikes occur, the second step of our pipeline transforms the time series into its backward
difference
\begin{align}
g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i, \label{eq:high-pass-filter}
\end{align}
as shown in Figure \ref{fig:hp_curve}.

To filter out small variations in the signal obtained after applying the
function $g$, as well as to eliminate negative values, we use the following
hard-threshold filter
\begin{align}\label{eqn:hfilter}
h(x^{t}_i) &= x^{t}_i \mathbb{1}(x^{t}_i \geq \tau) \text{ with } \tau > 0,
\end{align}
yielding Figure \ref{fig:threshold_curve} where $\tau$ is the threshold parameter and $\mathbb{1}$ is the indicator function.
As can be seen, the processed signal only contains clean spikes.

The objective of the last step of our filtering procedure is to decrease the
importance of spikes that occur when there is high global activity in the
network with respect to  spikes that occur during normal activity. Indeed, we
have conjectured that when a large part of the network is firing, the rate at
which observations are made is not high enough to be able to detect
interactions, and that it would therefore be preferable to lower their
importance by changing their magnitude appropriately. Additionally, it is
well-known that neurons may also spike because of a high global activity
\citep{stetter2012model}. In such  context, detecting pairwise neuron
interactions from the firing activity is meaningless. As such,
the signal output by $h$ is finally applied to the following function
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}}, \label{eq:magnify-filter}
\end{align}
whose effect is to magnify the importance of spikes that occur in cases of low
global activity (measured by $\sum_{j} x^{t}_j$), as observed, for instance,
around $t=4\text{s}$ in Figure~\ref{fig:weight_curve}. Note the particular case where there
is no activity, i.e., $\sum_{j} x^{t}_j = 0$, is solved by setting $w(x^{t}_i)
= 1$.

To summarise, the full signal processing pipeline of our simplified approach is defined by the composed function $w \circ h \circ g \circ
f_1$ (resp. $f_2$). When applied to the raw signal of Figure
\ref{fig:original_curve}, it outputs the signal shown in Figure
\ref{fig:weight_curve}.

@@

Our procedure to infer connections between neurons first assumes that
the (filtered) fluorescence concentrations of all $p$ neurons at each
time point can be modelled as a set of random variables $X = \{X_1,
\dots, X_p\}$ that are independently drawn from the same time-invariant
joint probability distribution $P_X$. 
As a consequence, our inference method does not exploit the time-
ordering of the observations (although time-ordering is exploited by
the filters).

Given this assumption, we then propose to use as a measure of the
strength of the connection between two neurons $i$ and $j$, the
\textit{Partial correlation} coefficient $p_{i,j}$ between their corresponding
random variables $X_i$ and $X_j$, defined by:
\begin{equation}
p_{i,j} =
-\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}, \label{eq:inverse}
\end{equation}
where $\Sigma^{-1}$, known as the precision or concentration matrix, is the inverse of the covariance matrix $\Sigma$ of $X$. 
Assuming that the distribution $P_X$ is a multivariate Gaussian
distribution ${\cal N}(\mu,\Sigma)$, it can be shown that $p_{i,j}$ is
zero if and only if $X_i$ and $X_j$ are independent given all other
variables in $X$, i.e., $X_i \perp X_j|X^{-i,j}$ where $X^{-i,j}= X
\setminus\{X_i,X_j\}$. Partial correlation thus measures conditional
dependencies between variables ; therefore it should naturally only detect direct associations
between neurons and filter out spurious indirect effects. The interest
of partial correlation as an association measure has already been
shown for the inference of gene regulatory networks
\citep{de2004discovery,Schafer:2005}.
Note that the partial correlation statistic is symmetric
(i.e. $p_{i,j}=p_{j,i}$). Therefore, our approach cannot identify the
direction of the interactions between neurons. We will see in
Section~\ref{sec:results} why this only slightly affects its
performance, with respect to the metric used in the Connectomics
Challenge.

Practically speaking, the computation of all $p_{i,j}$ coefficients using Equation
\ref{eq:inverse} requires the estimation of the covariance matrix $\Sigma$
and then computing its inverse. Given that typically we have more
samples than neurons, the covariance matrix can be inverted in a
straightforward way. We nevertheless obtained some improvement by
replacing the exact inverse with an approximation using only the $M$
first principal components \citep{bishop2006pattern} (with
$M=0.8 p$ in our experiments, see Appendix~\ref{app:pca}).

Finally, it should be noted that the performance of our simple method appears to
be quite sensitive to the values of parameters (e.g., choice of $f_1$ or $f_2$
or the value of the threshold $\tau$) in the combined function of the
filtering and inferring processes. One approach, further referred
to as \textit{Averaged Partial correlation} statistics, for improving
its robustness is to average correlation statistics over various
values of the parameters, thereby reducing the variance of its
predictions. Further details about parameter selection are provided in
Appendix~\ref{app:optimized}.

@@

We report here experiments on the \textit{normal-1,2,3}, and \textit{4}
datasets provided by the organisers of the Connectomics Challenge (see
Appendix~\ref{app:supp} for experiments on other datasets). Each of
these datasets is obtained from the simulation \citep{stetter2012model} of
different neural networks of 1,000 neurons and approximately 15,000 edges (i.e., a
network density of about 1.5\%). Each neuron is described by a calcium
fluorescence time-series of length $T=179500$. All inference methods compared
here provide a ranking of all pairs of neurons according to some association score. To assess the quality of this ranking, we compute both ROC and
precision-recall curves against the ground-truth network, which are represented by
the area under the curves and respectively  denoted AUROC and AUPRC. Only
the AUROC score was used to rank the challenge participants, but the precision-recall curve has been shown to be a more sensible metric for network
inference, especially when network density is small (see e.g.,
\cite{schrynemackers2013protocols}). Since neurons are not self-connected in
the ground-truth networks (i.e., $(i, i) \not \in E, \forall i \in V$), we
have manually set the score of such edges to the minimum possible association
score before computing ROC and PR curves.

@@

The top of Table \ref{tab:comparison} reports AUROC and AUPRC for all
four networks using, in each case, partial correlation with different
filtering functions. Except for the last two rows that use PCA, the
exact inverse of the covariance matrix was used in each case. These
results clearly show the importance of the filters. AUROC increases in
average from 0.77 to 0.93. PCA does not really affect AUROC scores, but
it significantly improves AUPRC scores. Taking the average over
various parameter settings gives an improvement of 10\% in AUPRC but
only a minor change in AUROC. The last row (``Full method'') shows the
final performance of the method specifically tuned for the challenge
(see Appendix \ref{app:optimized} for all details). Although this
tuning was decisive to obtain the best performance in the challenge,
it does not significantly improve either AUROC or AUPRC.

@@

At the bottom of Table \ref{tab:comparison}, we provide as a comparison the
performance of three other methods: standard (Pearson) correlation (PC),
generalised transfer entropy (GTE), and GENIE3. ROC and PR curves on the
\textit{normal-2} network are shown for all methods in Figure
\ref{fig:curves}. Pearson correlation measures the unconditional linear
(in)dependence between variables and it should thus not be able to filter out
indirect interactions between neurons. GTE \citep{stetter2012model} was
proposed as a baseline for the challenge. This method builds on Transfer
Entropy to measure the association between two neurons. Unlike our approach, it
can predict the direction of the edges. GENIE3 \citep{huynhthu2010inferring} is
a gene regulatory network inference method that was the best performer in the
DREAM5 challenge \citep{marbach2012}. When transposed to neural networks, this
method uses the importance score of variable $X_i$ in a Random Forest model trying to
predict $X_j$ from all variables in $X\setminus X_j$ as a confidence score for the edge going from neuron $i$ to neuron
$j$. However, to reduce the
computational cost of this method, we had to limit each tree in the
Random Forest model to a maximum depth of 3. This constraint has a potentially
severe effect on the performance of this method with respect to the use of
fully-grown trees. PC and GENIE3 were applied to the time-series filtered using the functions $w\circ h\circ g$ and $h\circ g\circ f_1$ (which
gave the best performance), respectively. For GENIE3, we built 10,000 trees per neuron and we
used default settings for all other parameters (except for the maximal tree
depth). For GTE, we reproduced the exact same setting (conditioning level and
pre-processing) that was used by the organisers of the challenge.

Partial correlation and averaged partial correlation clearly outperform all
other methods on all datasets (see Table \ref{tab:comparison} and Appendix \ref{app:supp}). The
improvement is more important in terms of AUPRC than in terms of AUROC. As
expected, Pearson correlation performs very poorly in terms of AUPRC. GTE and
GENIE3 work much better, but these two methods are nevertheless clearly below
partial correlation. Among these two methods, GTE is slightly better in terms
of AUROC, while GENIE3 is significantly better in terms of AUPRC. Given that we
had to limit this latter method for computational reasons, these results are
very promising and a comparison with the full GENIE3 approach is certainly part
of our future works.

The fact that our method is unable to predict edge directions does not seem to
be a disadvantage with respect to GTE and GENIE3. Although partial correlation
scores each edge, and its opposite, similarly, it can reach precision values
higher than 0.5 (see Figure \ref{fig:curves}(b)), suggesting that it mainly ranks high
pairs of neurons that interact in both directions.  It is interesting also to
note that, on \textit{normal-2}, a method that perfectly predicts the
undirected network (i.e., that gives a score of $1$ to each pair $(i,j)$ such that
$(i,j)\in E$ or $(j,i)\in E$, and $0$ otherwise) already reaches an AUROC as high
as $0.995$ and an AUPRC of $0.789$.

@@

In this paper, we outlined a simple but efficient methodology for the problem
of connectome inference from calcium imaging data. Our approach consists of two
steps: (i) processing fluorescence data to detect neural peak activities; and
(ii) inferring the degree of association between neurons from partial
correlation statistics. Its simplified variant outperforms other
network inference methods while its optimized version proved to be the best method
on the Connectomics Challenge. Given its simplicity and good performance, we
therefore believe that the methodology presented in this work
would constitute a solid and easily-reproducible baseline for further work in
the field of connectome inference.

\paragraph{Acknowledgments}
A. Joly and G. Louppe are research fellows of the FNRS, Belgium.  A. Sutera is a
recipient of an FRIA fellowship of FRS-FNRS, Belgium. This work is supported by
PASCAL2 and the IUAP DYSCO, initiated by the Belgian State, Science Policy
Office.

@@

This section provides a detailed description of the method specifically tuned
for the Connectomics Challenge. We restrict our description to the
differences with respect to the simplified method presented in the main
paper. Most parameters were tuned so as to maximize AUROC on the
\textit{normal-1} dataset and our design choices were validated by monitoring
the AUROC obtained by the 145 entries we submitted during the
challenge. Although the tuned method performs better than the simplified one on
the challenge dataset, we believe that the tuned method clearly overfits the
simulator used to generate the challenge data and that the simplified method
should work equally well on new independent datasets. We nevertheless provide
the tuned method here for reference purposes. Our implementation of the tuned
method is available at \url{https://github.com/asutera/kaggle-connectomics}.

@@

\begin{itemize}
\item In addition to $f_1$ and $f_2$ (Equations \ref{eq:symetric-median} and
  \ref{eq:weighted-asymetric-median}), two alternative low-pass filters $f_3$
  and $f_4$ are considered:
\begin{align}
f_3(x^t_i) &= x^{t-1}_i + x^{t}_i + x^{t+1}_i + x^{t+2}_i, \label{eq:asymetric-median-forward} \\
f_4(x^t_i) &=  x_i^t + x^{t+1}_i  + x^{t+2}_i + x^{t+3}_i. \label{eq:asymetric-median}
\end{align}
\item An additional filter $r$ is applied to smoothe differences in peak magnitudes
  that might remain after the application of the hard-threshold filter $h$:
\begin{align}
r(x^t_i) = (x_i^t)^c,
\end{align}
with $c=0.9$.
\item Filter $w$ is replaced by a more complex filter $w^*$ defined as:
\begin{align}
 w^*(x^{t}_i) &= {(x^{t}_i + 1 )^{\left (1 + \frac{1}{\sum_{j} x^{t}_j}\right )}}^{k(\sum_{j} x^{t}_j)}
\end{align}
where the function $k$ is a piecewise linear function optimised separately for
each filter $f_1$, $f_2$, $f_3$ and $f_4$ (see the implementation for full
details). Filter $w$ in the simplified method is a special case of $w^*$ with
$k(\sum_j x_j^t)=1$.
\end{itemize}
The pre-processed time-series are then obtained by the application of the
following function: $w^*\circ r \circ h \circ g \circ f_i$ (with $i=1$, 2, 3, or 4).

@@

As discussed in Section \ref{sec:inference}, the performance of the method (in
terms of AUROC) is sensitive to the value of the parameter $\tau$ of the
hard-threshold filter $h$ (see Equation \ref{eqn:hfilter}), and to the choice
of the low-pass filter (among $\{f_1, f_2, f_3, f_4\}$).
As in the simplified method, we have averaged the partial correlation statistics obtained for all the pairs $(\tau,\mbox{low-pass filter}) \in \{0.100,0.101,\ldots,0.209\}\times \{f_1, f_2, f_3, f_4\}$.

Filters $f_1$ and $f_2$ display similar performances and thus were given similar
weights (i.e., resp. $0.383$ and $0.345$). These weights were chosen equal to the weights selected for the simplified method. In contrast, filters $f_3$
and $f_4$ turn out, individually, to be less competitive and were therefore given
less importance in the weighted average (i.e., resp. $0.004$ and $0.268$). Yet, as further shown in
Section~\ref{sapp:results}, combining all $4$ filters proves to marginally
improve performance with respect to using only $f_1$ and $f_2$.
